Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2025-05-19 manuscript/manuscript.tex
Include Flowchart & Update Fig5

Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
\end{itemize}
If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
Additionally, the MD simulation files and raw GIST output files can be found at \url{https://doi.org/10.48323/4mbrd-67m83}, to reproduce the results shown in this tutorial.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
For help with the listed cpptraj commands, AMBER atom mask syntax and other AMBER-specific commands, we recommend the following webresources:\\
\begin{itemize}
\item The website https://amberhub.chpc.utah.edu/ which is managed by the Cheatham III group, 
\item The cpptraj manual at https://raw.githubusercontent.com/Amber-MD/cpptraj/master/doc/CpptrajManual.pdf 
\item and the AMBER manual \cite{amber24} at https://ambermd.org/Manuals.php.
\end{itemize}
\pagebreak %TODO: pagebreak, maybe remove these by better formatting

@@

\subsection{Tutorial data}
We will use the crystal structure of streptavidin (PDB code: 1STP \cite{Weber1989-streptavidin-structure}) to start our calculations.
For section \ref{subsec:gist_biotin}, we will use the biotin ligand extracted from the crystal structure.
For the second part in section \ref{subsec:gist_complex}, we will use the whole crystal structure.
For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
\begin{lstlisting}[style=bash]
git clone git@github.com:liedllab/gist-tutorial.git
\end{lstlisting}
This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.
Additionally, the original MD trajectory files and raw GIST output files are provided under the DOI \url{https://doi.org/10.48323/4mbrd-67m83}.
\subsection{Running GIST for Biotin}
\label{subsec:gist_biotin}
In this example, we will be running a simple GIST calculation of the biotin solute molecule to get familiar with the usage of GIST in \software{cpptraj}.
\subsubsection{System preparation}
First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
The solute molecule should be solvated with a sufficiently large solvent box. 
For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images. 
This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water \cite{Chen2021,Waibl2022-gist-solvents} though we recommend \SI{15}{\angstrom} or higher.
Carefully check the employed force field, protonation state and partial charges.
%a PDB file containing only biotin. Prepare your structure by adding hydrogens, making sure to choose the correct protonation state. Parametrize biotin with the forcefield of your choice. Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
A good starting point for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).

@@

\subsubsection{Visualizing $\dgsolv$}
Next, we visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx('A\_dens', 'A\_dens.dx')}.
It might also be interesting to visualize the average energy of a water solvent molecule at each grid voxel.
This quantity is given by $2E_{ww} + E_{sw}$.
The energy referencing leads to non-zero normalized values where the population is zero.
While this is irrelevant for further post-processing, it is advantageous for visualization to set those empty regions to zero.
An OpenDX file can be produced using \software{gisttools} as follows:

\begin{lstlisting}[style=python]
import gisttools as gt
gist = gt.gist.load_gist_file('gist.dat',
    struct='solute-centered.pdb', eww_ref=-9.533)
gist['E_norm'] = gist['Eww_norm'] * 2 + gist['Esw_norm']
gist.loc[gist['population'] == 0, 'E_norm'] = 0
gist.save_dx('E_norm', 'gist-E-per-mol-norm.dx')
\end{lstlisting}
The OpenDX files can then be imported into your molecular viewer of choice.
A visualization of entropy and energy contributions to the free energy of solvation of the streptavidin binding pocket is shown in Figure~\ref{fig_binding_pocket_pymol}. 
PyMOL input scripts to generate such visualizations are available in the GitHub repository for this tutorial.

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/binding_pocket_norm_quants.png}
	\caption{Free energy contributions in the streptavidin binding pocket. Red: high solvent energy (at +2.5 
	kcal/mol/molecule). Green: low solvent energy (at -2.5 kcal/mol/molecule). Blue: low solvent entropy $T\Delta S^{six}$ 
	(at -2.5 kcal/mol/molecule).}\label{fig_binding_pocket_pymol}
\end{figure}

@@

\section{Checklists and Cheat Sheats} \label{sec:checklists}
%Tutorials do not necessarily require the use of a checklist as in Best Practices documents; however, they can include these if desired.
%Several useful checklist formats are available, with examples presented in \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates}.
%One example is shown here.

@@

\end{tabularx}
\end{checklist}
\end{Checklists*}
\FloatBarrier
\section*{Author Contributions}

>>>>

Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
\end{itemize}
If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
For help with the listed cpptraj commands, AMBER atom mask syntax and other AMBER-specific commands, we recommend the following webresources:\\
\begin{itemize}
\item The website https://amberhub.chpc.utah.edu/ which is managed by the Cheatham III group, 
\item The cpptraj manual at https://raw.githubusercontent.com/Amber-MD/cpptraj/master/doc/CpptrajManual.pdf 
\item and the AMBER manual \cite{amber24} at https://ambermd.org/Manuals.php.
\end{itemize}
\pagebreak %TODO: pagebreak, maybe remove these by better formatting

@@

\subsection{Tutorial data}
We will use the crystal structure of streptavidin (PDB code: 1STP \cite{Weber1989-streptavidin-structure}) to start our calculations.
For section \ref{subsec:gist_biotin}, we will use the biotin ligand extracted from the crystal structure.
For the second part in section \ref{subsec:gist_complex}, we will use the whole crystal structure.
For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
\begin{lstlisting}[style=bash]
git clone git@github.com:liedllab/gist-tutorial.git
\end{lstlisting}
This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.
Additionally, a link to download the original MD trajectory files is provided in the GitHub repository.
\subsection{Running GIST for Biotin}
\label{subsec:gist_biotin}
In this example, we will be running a simple GIST calculation of the biotin solute molecule to get familiar with the usage of GIST in \software{cpptraj}.
\subsubsection{System preparation}
First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
The solute molecule should be solvated with a sufficiently large solvent box. 
For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images. 
This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water \cite{Chen2021,Waibl2022-gist-solvents} though we recommend \SI{15}{\angstrom} or higher.
Carefully check the employed force field, protonation state and partial charges.
%a PDB file containing only biotin. Prepare your structure by adding hydrogens, making sure to choose the correct protonation state. Parametrize biotin with the forcefield of your choice. Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
A good starting point for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).

@@

\subsubsection{Visualizing $\dgsolv$}
Next, we visualize the enthalpic contributions to the free energy of solvation.
You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx('A\_dens', 'A\_dens.dx')}.
It might also be interesting to visualize the average energy of a water solvent molecule at each grid voxel.
This quantity is given by $2E_{ww} + E_{sw}$.
The energy referencing leads to non-zero normalized values where the population is zero.
While this is irrelevant for further post-processing, it is advantageous for visualization to set those empty regions to zero.
An OpenDX file can be produced using \software{gisttools} as follows:

\begin{lstlisting}[style=python]
import gisttools as gt
# Load the GIST data, make sure to set the correct reference energy
gist = gt.gist.load_gist_file('gist.dat',
    struct='solute-centered.pdb', eww_ref=-9.533)
gist['E_norm'] = gist['Eww_norm'] * 2 + gist['Esw_norm']
gist.loc[gist['population'] == 0, 'E_norm'] = 0
gist.save_dx('E_norm', 'gist-E-per-mol-norm.dx')
\end{lstlisting}
The OpenDX files can then be imported into your molecular viewer of choice.
A visualization of enthalpy contributions to the free energy of solvation of the streptavidin binding pocket is shown in Figure~\ref{fig_binding_pocket_pymol}. 
Enthalpywise, the streptavidin binding pocket is notably hydrophobic (red) in most parts except for where biotin is also interacting polarly with the protein (green).
Nothe that this is "norm" data, so not density-weighted but rather comparing voxels on a per-molecule basis.
The surfaces are therefore disjointed since even voxels seldomly sampled by the water are shown.
PyMOL input scripts to generate such visualizations are available in the GitHub repository for this tutorial.

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/binding_pocket_norm_quants.png}
	\caption{Enthalpy contributions using "norm" data in the streptavidin binding pocket overlayed with biotin. Red: high solvent energy (at +3 
	kcal/mol/molecule). Green: low solvent energy (at -3 kcal/mol/molecule). The enthalpy density is based on the apo structure and is shown only when within 1.5\AA of biotin.}\label{fig_binding_pocket_pymol}
\end{figure}

@@

\section{Checklists, Cheat Sheats ad Flowcharts} \label{sec:checklists}
%Tutorials do not necessarily require the use of a checklist as in Best Practices documents; however, they can include these if desired.
%Several useful checklist formats are available, with examples presented in \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates}.
%One example is shown here.

@@

\end{tabularx}
\end{checklist}
\end{Checklists*}

\begin{Checklists*}
	\includegraphics[width=0.9\linewidth,page=1]{figures/flowchart_GIST.pdf}
\end{Checklists*}
	\pagebreak
	\begin{Checklists*}
	\includegraphics[width=0.9\linewidth,page=2]{figures/flowchart_GIST.pdf}
\end{Checklists*}
\FloatBarrier

\section*{Author Contributions}
